Distributional Trees for Circular Data

### 2019-04-09 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
note on scale of wind direction and correction of typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2523 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 11b45c1d109939aab524a814cff592115f89740f

Circular data can be found in a variety of applications and subject areas, 
e.g., hourly crime rate in the social-economics, 
animal movement direction or gene-structure in biology, 
and wind direction as one of the most important weather variables in meteorology.
Circular regression models were first introduced by Fisher and Lee (1992) and 
further extended by Jammalamadaka and Sengupta (2001) and 
Mulder and Klugkist (2017) among others.
While most of the already existing approaches are built on additive regression models, 
we propose an adaption/extension of regression trees to circular data by employing 
distributional trees.

@@

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision-making processes and risk management,
e.g., in air traffic management or renewable energy production. This study
employs circular regression trees to obtain hourly wind direction
forecasts at Innsbruck Airport.
Innsbruck lies at the bottom of a deep valley in the Alps. Topography
channels wind along the west-east valley axis or along a tributary valley
intersecting from the south. Hence, pressure gradients to which valley wind
regimes react both west and east of the airport are considered as covariates
along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

@@

\section{Discussion and Outlook}

@@

\subsection{Ensembles and random forests}
A natural extension are ensembles or forests of such circular trees
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by Breiman~et.~al~(1984) average the
predictions of an ensemble of trees, each built on a subsample 
or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions
by adaptive local likelihood estimation of the distributional parameters
(Schlosser~et.~al, 2019). More specificially, for each possibly new
observation $x$ a set of ``nearest neighbor'' weights $w_i(x)$ is obtained 
that is based on how often $x$ is assigned to the same terminal node as
each learning observation $y_i, i \in \{1,\ldots,n\}$.

The parameters $\mu$ and $\kappa$ are then estimated for each (new) 
observation $x$ by weighted maximum likelihood estimator based on the
adaptive nearest neighbor weights:
\begin{equation}
\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}
Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $x$ whereas $w_i(x) = 1$ would correspond to the unweighted
full-sample estimates and $w_i(x) \in \{0, 1\}$ corresponds to the abrupt
splits from the tree.

@@

In order to obtain more parsimonious and more stable trees another possible
extension for \emph{circular covariates} (with or without a \emph{circular response})
is to consider their circular nature when searching the best split into two segments.
In general, searching the best separation of a covariate into a ``left'' and ``right''
daugther node tries to maximize the segmented log-likelihood:
\begin{equation}
\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
\end{equation}
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a circular covariate $\in [0, 2 \pi)$
only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$, encompassing zero. The latter
strategy is invariant to the (often arbitrary) definition of the direction at zero.

@@

The required test statistic to maximally select two split points simultaneously is straightforward
to accomodate in the CTree framework by providing all binary indicators corresponding
to the splints into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
increasingly slow for larger sample sizes but it might be possible to speed up computations by
exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
MOB framework the extension is not quite as straightforward but one strategy could be to
adapt double maximum tests \`a la Bai and Perron~(2003).

@@

\subsubsection*{Computational Details:}
\textsf{R} packages implementing the proposed methods are currently under development at
\texttt{https://R-Forge.R-project.org/} \texttt{projects/partykit/}.

>>>>

Circular data can be found in a variety of applications and subject areas, 
e.g., hourly crime rate in the social-economics, 
animal movement direction or gene-structure in biology, 
and wind direction as one of the most important weather variables in meteorology.
Circular regression models were first introduced by Fisher and Lee (1992) and 
further extended by Jammalamadaka and Sengupta (2001) and 
Mulder and Klugkist (2017) among others.
While most of the already existing approaches are built on additive regression models, 
we propose an adaption of regression trees to circular data by employing 
distributional trees.

@@

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision-making processes and risk management,
e.g., in air traffic management or renewable energy production. This study
employs circular regression trees to obtain hourly wind direction
forecasts at Innsbruck Airport.
Innsbruck lies at the bottom of a deep valley in the Alps. Topography
channels wind along the west-east valley axis or along a tributary valley
intersecting from the south. Hence, pressure gradients to which valley wind
regimes react both west and east of the airport are considered as covariates
along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.
Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).

@@

\section{Discussion and outlook}

@@

\subsection{Ensembles and random forests}
A natural extension are ensembles or forests of such circular trees
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by Breiman~et.~al~(1984) average the
predictions of an ensemble of trees, each built on a subsample 
or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions
by adaptive local likelihood estimation of the distributional parameters
(Schlosser~et.~al, 2019). More specifically, for each possibly new
observation~$x$ a set of ``nearest neighbor'' weights $w_i(x)$ is obtained 
that is based on how often $x$ is assigned to the same terminal node as
each learning observation $y_i, i \in \{1,\ldots,n\}$.

The parameters $\mu$ and $\kappa$ are then estimated for each (new) 
observation $x$ by weighted maximum likelihood based on the
adaptive nearest neighbor weights:
\begin{equation}
\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}
Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $x$ whereas $w_i(x) = 1$ would correspond to the unweighted
full-sample estimates and $w_i(x) \in \{0, 1\}$ corresponds to the abrupt
splits from the tree.

@@

In order to obtain more parsimonious and more stable trees another possible
extension for \emph{circular covariates} (with or without a \emph{circular response})
is to consider their circular nature when searching the best split into two segments.
In general, searching the best separation of a covariate into a ``left'' and ``right''
daughter node tries to maximize the segmented log-likelihood:
\begin{equation}
\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
\end{equation}
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a circular covariate $\in [0, 2 \pi)$
only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$, encompassing zero. The latter
strategy is invariant to the (often arbitrary) definition of the direction at zero.

@@

The required test statistic to maximally select two split points simultaneously is straightforward
to accommodate in the CTree framework by providing all binary indicators corresponding
to the splits into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
increasingly slow for larger sample sizes but it might be possible to speed up computations by
exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
MOB framework the extension is not quite as straightforward but one strategy could be to
adapt double maximum tests \`a la Bai and Perron~(2003).

@@

\subsubsection*{Computational details:}
\textsf{R} packages implementing the proposed methods are currently under development at
\texttt{https://R-Forge.R-project.org/} \texttt{projects/partykit/}.
